Ages of Planets and Stars Observed by Kepler

### 2024-04-20 paper/ms.tex
update ms

%TODO: plot uncertainties in teff vs teff...
%TODO: plot uncertainties in logg vs logg...

%TODO: add that plot showing differences btwn the two teff scales?

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}.  We built this statistical framework to address
the issue that, especially over the first gigayear of stellar
spin-down, stars with the same mass and same rotation period can have
very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
gyrochrone age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
measured.  The three main assumptions behind the \texttt{gyro-interp}
framework are as follows.  {\it 1)} The stars are on the main
sequence, {\it 2)} their spindown is not influenced by binary
companions, and {\it 3)} the rotation period distribution for field
stars of a given mass and age is identical to that of the calibration
open clusters (metallicity differences, for instance, are ignored).  
Regarding the former two assumptions: pre-main-sequence and
post-main-sequence evolution introduce variations in the stellar
angular momentum budget.  Similarly, tidal effects might
systematically spin up (or down) stars in binaries.

Using our adopted effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is only calibrated
out to 4\,Gyr, the adopted age of the M67 cluster
\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the stellar
spin-down rate as a function of time and stellar temperature, this
framework interpolates between reference open clusters using smooth
cubic hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citep{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf The models.}
		Points represent $10^4$ draws from model probability distributions
		that have been fitted to rotation periods 
		\citep{Bouma_2023} and to lithium measurements \citep{Jeffries_2023}
		of stars in open clusters.
		Lines are the fitted ``mean models'' at various ages; these models are calibrated
		for stars younger than 4\,Gyr.
		The intrinsic dispersion in rotation and lithium at any given age, which what these models are designed to fit, sets
		the theoretical precision limit for both age-dating methods.
		Additional sources of uncertainty, including measurement uncertainty,
		impose additional limitations on achievable age precision.
		The displayed points assume an unphysical uniform distribution in
		temperature for visual clarity, and sizes of the points are the same in each panel.
		%    Gyrochrone age posteriors are broadest in regions 
		%    with the slowest stellar spin-down.
		\label{fig:models}
	}
\end{figure*}

\subsection{Gyrochronology Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select stars for which we expect these age to in fact be valid, we
then build a set of quality flags.  When and how these flags should be
applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they might all be applied.  If
the goal is constructing a complete sample of young stars, then
consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  The flags in
Tables~\ref{tab:stars} and~\ref{tab:planets} enable selecting for or
against such cases using bit flags.

{\it Temperature range (bit 0)}---We require stars to have effective
temperatures $T_{\rm eff}$ between 3800 and 6200\,K in order to report
gyrochrone ages \citep{Bouma_2023}.   Stars hotter than 6200\,K spin
down very slowly, if at all.  Stars cooler than 3800\,K do spin down
over gigayear timescales (e.g. {\bf CITE} E.Newton thesis, the Engle
paper, and Chiti's article), but the currently available open cluster
data have yet to clarify at what times, if any, the intrinsic scatter
in the population decreases.

>>>>

% optional todo's:
%TODO: plot uncertainties in teff vs teff...
%TODO: plot uncertainties in logg vs logg...
%TODO: add that plot showing differences btwn the two teff scales?

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}.  We designed this method to address
the issue that, especially over the first gigayear of stellar
spin-down, stars with the same mass and same rotation period can have
very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
gyrochrone age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
inferred.  The three main assumptions behind the \texttt{gyro-interp}
framework are as follows.  {\it 1)} The stars are near the main
sequence, {\it 2)} their spindown is not influenced by binary
companions, and {\it 3)} the rotation period distribution for field
stars of a given mass and age is identical to that of equally aged
open clusters (metallicity differences, for instance, are ignored).  
Regarding the former two assumptions: pre-main-sequence contraction and
post-main-sequence expansion influence surface stellar rotation periods
in ways that we do not model.  Stars in e.g.~ close binaries might be
subject to tidal effects that we do not treat.

Using our adopted effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is calibrated
out to 4\,Gyr, the adopted age of the M67 cluster
\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the stellar
spin-down rate as a function of time and stellar temperature, this
framework interpolates between reference open clusters using smooth
cubic hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citet{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf The models.}
		Points represent $10^4$ draws from model probability distributions
		that have been fitted to observed rotation periods 
		\citep{Bouma_2023} and lithium equivalent widths \citep[EWs;][]{Jeffries_2023}
		of stars in open clusters.
		Lines are the ``mean models'' at various ages;
		the asymmetric intrinsic dispersion in rotation and lithium about these mean models, which is what the models fit, sets
the theoretical precision limit for both age-dating methods.		
		Additional sources of uncertainty, including measurement uncertainty,
		impose further limits on achievable age precision.
		These models are calibrated
		using open clusters younger than 4\,Gyr.
		The displayed points assume an unphysical uniform distribution in
		temperature for visual clarity.
		The sizes of the points are the same in each panel, so that
		low apparent density signifies greater dispersion around the mean.
		\label{fig:models}
	}
\end{figure*}

\subsection{Gyrochronology Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
flag stars for which we suspect these ages may not in fact be valid,
we then build a set of bit-quality flags.  When and how these flags should
be applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they might all be applied.  If
the goal is constructing a complete sample of young stars, then
consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  Our quality flags enable
selecting for or against such cases by selecting from the following
bits.

{\it Temperature range (bit 0)}---We require stars to have effective
temperatures $T_{\rm eff}$ between 3800 and 6200\,K in order to report
gyrochrone ages \citep{Bouma_2023}.   Stars hotter than 6200\,K spin
down very slowly, if at all.  Stars cooler than 3800\,K do spin down
over gigayear timescales
\citep{2016ApJ...821...93N,2023ApJ...954L..50E,2024arXiv240312129C},
but the currently available open cluster data have yet to clarify
exactly when the intrinsic scatter in the population decreases.
